---
title: "613HW4"
author: "Yifan Wang"
date: "2022/4/8"
output: html_document
---

```{r setup, include=FALSE}
library(dplyr)
library(tidyr)
library(ggplot2)
library(censReg)
library(fastDummies)
library(sampleSelection)
library(panelr)
```
```{r }
dat_A4 <- read.csv("~/Desktop/Data/dat_A4.csv")
dat_A4_panel <- read.csv("~/Desktop/Data/dat_A4_panel.csv")
panel=dat_A4_panel[,-1]
dat_A4=dat_A4%>%rename(income=YINC_1700_2019)%>%rename(sex=KEY_SEX_1997)%>%rename(nchildren=CV_BIO_CHILD_HH_U18_2019)%>%rename(marital=CV_MARSTAT_COLLAPSED_2019)%>%rename(edudegree=YSCH.3113_2019)%>%rename(race=KEY_RACE_ETHNICITY_1997)
```
#EX1

***1) Create additional variable age and total work experience measured in years***
```{r }
dat4= dat_A4 %>% mutate(age=2022-KEY_BDATE_Y_1997) %>%
    rowwise() %>% 
    mutate(work_exp = sum(c_across(contains("JOB"))*7/365, na.rm = T))
```
***2) Create additional education variables indicating total years of schooling***
```{r }
#recode individual's highest degree: 4 years for GED; 2 years for AA; 4 years for BA; 2 years for MA; 5 years for Phd; 6 years higher edu for Professional degree
dat4 = dat4 %>% mutate(totaledu = as.numeric(recode(edudegree, '1' = "0",'2' = "4", '3' = "12",'4' = "14",'5' = "16",'6' = "18",'7' = "23",'8' = "18")))

#since parents' education already in total years of schooling, so only filter out ungraded (95)
dat4 = dat4 %>% filter(CV_HGC_BIO_DAD_1997!="95" & CV_HGC_BIO_MOM_1997!="95" & CV_HGC_RES_DAD_1997!="95" & CV_HGC_RES_MOM_1997!="95")
```
***3) Visualizations***
```{r }
#i)Plot the income data by age groups
ggplot(dat4, aes(x = factor(age), y = income, na.rm = T)) +
   geom_boxplot(outlier.colour=NA) +
   coord_cartesian(ylim = c(0, 100000))+ labs(title= "Boxplot of income distribution by age groups")
```
```{r }
#ii)Plot the income data by gender groups
ggplot(dat4, aes(x = factor(sex), y = income,na.rm = T)) +
   geom_boxplot(outlier.colour=NA) +
   coord_cartesian(ylim = c(0, 100000))+ labs(title= "Boxplot of income distribution by gender groups")
```
```{r }
#iii)Plot the income data by number of children
ggplot(dat4, aes(x = factor(nchildren), y = income,na.rm = T)) +
   geom_boxplot(outlier.colour=NA) +
   coord_cartesian(ylim = c(0, 100000))+ labs(title= "Boxplot of income distribution by number of children")
```
```{r }
#i)Table the share of "0" in the income data by age groups
income0=dat4 %>% filter(income == 0)
table1 = as.table(table(income0$age))
prop.table(table1)
```
```{r }
#ii)Table the share of ”0” in the income data by gender groups
table2 = as.table(table(income0$sex))
prop.table(table2)
```
```{r }
#iii)Table the share of ”0” in the income data by number of children and marital status
#by number of children (separated)
table3 = as.table(table(income0$nchildren))
prop.table(table3)
#by marital status (separated)
table4 = as.table(table(income0$marital))
prop.table(table4)
#by number of children and marital status (combined)
table5 = as.table(table(income0$marital,income0$nchildren))
prop.table(table5)
```

***Interpretation of Visualization***

***i)Boxplot of income by age groups shows that there is no significant difference between different age groups***

***ii)Boxplot of income by gender groups shows that male have higher average income than female***

***iii)Boxplot of income by number of children shows that people with 1 to 3 children have higher average income than others***

***i)Table of share of no income by age groups shows that people who are age 38 and 41 have larger proportion than other age groups***

***ii)Table of share of no income by gender groups shows that male have higher proportion than female***

***iii)Table of share of no income by number of children shows that people with 0 to 2 children have higher proportion than others;Table of share of no income by marital status shows that people who are never married and married have higher proportion than others;Table of share of no income by both number of children and marital status shows that for people who are married having additional child increase  the proportion of no income***

#EX2

***1)Specify and estimate an OLS model to explain the income variable***
```{r }
dat4$marital=as.factor(dat4$marital)
dat4$race=as.factor(dat4$race)
dat4$sex=as.factor(dat4$sex)
ex2_model1 = dat4 %>% filter(income>0)%>%select(income,age,work_exp,totaledu,nchildren,marital,sex,race) %>% na.omit() %>% mutate(ln_income=log(income))
model1=lm(ln_income~ age + work_exp + totaledu, data = ex2_model1)
summary(model1)
```

***1a)Interpretation of OLS estimates***
***age: for each additional year of age, income are predicted to increase by 0.36%***
***work_exp: for each additional year of work experience, income are predicted to increase by 3.9%***
***totaledu:: for each additional year of schooling, income are predicted to increase by 5.2%***

***1b)Selection problem: Part of dependent variable is unobserved (there are 36 unemployed people, but has wage observations for only those who work) Since people who work are selected non-randomly from the population, estimating the determinants of wages from the subpopulation who work may introduce bias.***

***EX2 Q2)Heckman model: in the first stage, run a probit model to predict employment probability for each individual;In the second stage,by incorporating a transformation of these predicted individual probabilities as an additional explanatory variable, it corrects for self-selection.***

***EX2 Q3)***
```{r }
#3)Estimate a Heckman selection model
ex2_model2 =dat4 %>%select(income,age,work_exp,totaledu,nchildren,marital,sex,race) %>%
  na.omit() %>% 
  mutate(ln_income=log(income+1)) %>%  mutate(employed=case_when(income>0~1,TRUE~0))
#Probit by glm
reg1 = glm(employed~ age + work_exp + totaledu + nchildren + marital + sex + race,family = binomial(link = "probit"), data=ex2_model2)
par=as.vector(summary(reg1)$coefficients[, 1])
#find IMR
xbeta <- as.matrix(predict(reg1))
ex2_model2$imr=dnorm(xbeta)/pnorm(xbeta)
ex2_model2_obs=ex2_model2 %>% filter(income>0) #select positive income
```
```{r }
#additionally add imr as a regressor to OLS
model2=lm(ln_income~ age + work_exp + totaledu  + imr, data = ex2_model2_obs)
summary(model2)
```
```{r }
#Verify result with sampleSlection package
heckit=selection(employed~ age + work_exp + totaledu + nchildren + marital + sex + race,
                 ln_income~ age + work_exp + totaledu,
                 method = "2step",
                 data = ex2_model2)
summary(heckit) #same estimates
```

***Interpretation for Heckman model***
***age: for each additional year of age, income are predicted to increse by 0.11%***
***work_exp: for each additional year of work experience, income are predicted to increse by 3.8%***
***totaledu:: for each additional year of schooling, income are predicted to increse by 5.1%***
***A negative coefficient for the inverse Mills ratio in the OLS regression means that observed wages are less on average than offer wages***

```{r }
#compare to OLS
OLS=rbind(as.matrix(summary(model1)$coefficients[, 1]),0)
compareOLS = cbind(summary(model2)$coefficients[, 1],OLS)
colnames(compareOLS) = c("Heckman est","OLS est")
compareOLS 
```
***There exist a difference because people who are unemployed are now included in the sample***

#EX3
***1)Plot a histogram to check whether the distribution of the income variable***
```{r }
ggplot(dat4,aes(x=income))+geom_histogram()
```

***The income variable is right censored at 100000***

***2) I propose the Tobit model (regress ln(income) on age, work experience and education)***
```{r }
tobitmodel= censReg(ln_income ~ age + work_exp + totaledu , data = ex2_model1, right = log(100000))
tobit=summary(tobitmodel)
tobit
```
***3) likelihood function for Tobit***
```{r }
#3) likelihood function for Tobit
tobitlike <- function(par, X, yvar){
  xbeta <- X %*% par[1:ncol(X)]
  sigma <- par[ncol(X)+1]
  d = ifelse(yvar>=log(100000),0,1) #indicator
  pr <- pnorm((log(100000)-xbeta)/sigma)
  pdf<- dnorm((yvar-xbeta)/sigma)
  pr[pr>0.999999] = 0.999999
  pr[pr<0.000001] = 0.000001
  like = d*log(1/sigma*pdf) + (1-d)*log(1-pr)
  return(-sum(like)) 
}
# X, y (data) for exercise 3
df_ex3 <- ex2_model1 %>% mutate(intercept = 1) %>%
  select(ln_income,intercept,age,work_exp,totaledu)
X = as.matrix(df_ex3%>%select(-ln_income))
y = as.matrix(df_ex3%>%select(ln_income))
#optimize coeficient
set.seed(4)
par = runif(ncol(X)+1)
res2 = optim(par,fn=tobitlike,method="BFGS",control=list(trace=6,REPORT=1,maxit=1000),X=X,yvar=y,hessian=TRUE)
res2$par[5]=log(res2$par[5])#logSigma
comparetobit = cbind(tobit$estimate[,1],res2$par)
colnames(comparetobit) = c("censReg est","own tobit est")
comparetobit 
```

***4)Interpretation of Tobit estimates***

***age: for each additional year of age, the predicted value of income increase by 1%***

***work_exp: for each additional year of work experience, the predicted value of income increase by 4.3%***

***totaledu:: for each additional year of schooling, the predicted value of income increase by 58%***

```{r }
#compare to not correcting for censored data
comparetobit2 = cbind(res2$par[1:5],OLS)
colnames(comparetobit2) = c("Tobit est","OLS est")
rownames(comparetobit2) = c("(Intercept)","age","work_exp","totaledu","logSigma")
comparetobit2
```

#EX4

***1)Explain the potential ability bias when trying to explain to understand the determinants of wages***

***There is a causal relationship between education and income. Education is a source of an accumulation of competences, and most productive individuals have an interest in studying for the longest period, entailing the probability of ability bias.***

***2)Exploit the panel dimension of the data to propose a model to correct for the ability bias***.
```{r }
#change inconsistent variable names for converting to long
panel <- panel %>% rename(CV_HIGHEST_DEGREE_EVER_EDT_1998=CV_HIGHEST_DEGREE_9899_1998)%>% rename(CV_HIGHEST_DEGREE_EVER_EDT_1999=CV_HIGHEST_DEGREE_9900_1999)%>% rename(CV_HIGHEST_DEGREE_EVER_EDT_2000=CV_HIGHEST_DEGREE_0001_2000)%>% rename(CV_HIGHEST_DEGREE_EVER_EDT_2001=CV_HIGHEST_DEGREE_0102_2001)%>% rename(CV_HIGHEST_DEGREE_EVER_EDT_2002=CV_HIGHEST_DEGREE_0203_2002)%>% rename(CV_HIGHEST_DEGREE_EVER_EDT_2003=CV_HIGHEST_DEGREE_0304_2003)%>% rename(CV_HIGHEST_DEGREE_EVER_EDT_2004=CV_HIGHEST_DEGREE_0405_2004)%>% rename(CV_HIGHEST_DEGREE_EVER_EDT_2005=CV_HIGHEST_DEGREE_0506_2005)%>% rename(CV_HIGHEST_DEGREE_EVER_EDT_2006=CV_HIGHEST_DEGREE_0607_2006)%>% rename(CV_HIGHEST_DEGREE_EVER_EDT_2007=CV_HIGHEST_DEGREE_0708_2007)%>% rename(CV_HIGHEST_DEGREE_EVER_EDT_2008=CV_HIGHEST_DEGREE_0809_2008)%>% rename(CV_HIGHEST_DEGREE_EVER_EDT_2009=CV_HIGHEST_DEGREE_0910_2009)
#convert wide to long
long <- long_panel(panel,prefix='_',begin  = 1997, end = 2019,label_location = "end")
#change variable names
long <-long %>% rename(year=wave) %>%rename(income=YINC.1700) %>% rename(marital=CV_MARSTAT_COLLAPSED)%>%rename(edu=CV_HIGHEST_DEGREE_EVER_EDT) 
```
```{r }
#generate total work experience
long2 <-long %>% rowwise() %>% 
    mutate(work_exp = sum(c_across(contains("JOB"))*7/365, na.rm = T))%>%
    select(id,year,income,edu,marital,work_exp)%>%na.omit()

#recode individual's highest degree: 4 years for GED; 2 years for AA; 4 years for BA; 2 years for MA; 5 years for Phd; 6 years higher edu for Professional degree
long2 = long2 %>% mutate(edu = as.numeric(recode(edu,'0' = "0",'1' = "4", '2' = "12",'3' = "14",'4' = "16",'5' = "18",'6' = "23",'7' = "18")))
```

***Within Estimator***
```{r }
income_i_bar= long2 %>% group_by(id) %>% summarise(income_i_bar=mean(income))
edu_i_bar= long2 %>% group_by(id) %>% summarise(edu_i_bar=mean(edu))
marital_i_bar= long2 %>% group_by(id) %>% summarise(marital_i_bar=mean(marital))
workexp_i_bar= long2 %>% group_by(id) %>% summarise(workexp_i_bar=mean(work_exp))

within= long2 %>% left_join(income_i_bar,by="id")%>% left_join(edu_i_bar,by="id") %>% left_join(marital_i_bar,by="id") %>% left_join(workexp_i_bar,by="id")

within=within %>% group_by(id) %>% mutate(income_diff=income-income_i_bar) %>% mutate(edu_diff=edu-edu_i_bar) %>% mutate(marital_diff=marital-marital_i_bar) %>% mutate(workexp_diff=work_exp-workexp_i_bar)

reg1=lm(income_diff~edu_diff + marital_diff + workexp_diff, data= within)
summary(reg1)
```

***Interpretation for Within Estimator Model***

***edu: On average, for each additional year increase in education, income will increase 1848, controlling for all time-invariant heterogeneity ***

***workexp: On average, for each additional year increase in work experience, income will increase 2817, controlling for all time-invariant heterogeneity ***

***marital: a categorical variable, can not be interpreted***

***Between Estimator***
```{r }
between=income_i_bar %>% left_join(edu_i_bar,by="id") %>% left_join(marital_i_bar,by="id") %>% left_join(workexp_i_bar,by="id")

reg2=lm(income_i_bar~edu_i_bar + marital_i_bar + workexp_i_bar, data= between)
summary(reg2)
```

***Interpretation for Between Estimator Model***

***edu: On average, for each additional year increase in education, income will increase 1848, controlling for individual heterogeneity ***

***workexp: On average, for each additional year increase in work experience, income will increase 2817, controlling for individual heterogeneity ***

***marital: a categorical variable, can not be interpreted***

***First Difference Estimator***
```{r }
FD = long2 %>% group_by(id) %>% mutate(income_diff= income-lag(income)) %>% mutate(edu_diff= edu-lag(edu)) %>% mutate(marital_diff= marital-lag(marital)) %>% mutate(workexp_diff= work_exp-lag(work_exp))
reg3=lm(income_diff~ edu_diff + marital_diff + workexp_diff, data= FD)
summary(reg3)
```

***Interpretation for First Difference Estimator Model***

***edu: In year t-1, an increase of 1 year education will result in an average of 103 increase in new income.***

***workexp: In year t-1, an increase of 1 year work experience will result in an average of 1699 increase in new income.***

***marital: a categorical variable, can not be interpreted***

***EX3 Q3***

***Within estimator takes the time variation and discard the individual effects; Between estimator takes the individual effects and discard the time variation; First Difference estimator takes both effects***
